Bayesian inference of prehistoric population dynamics from multiple proxies: a case study from the North of the Swiss Alps

### 2022-05-30 manuscript/paper_extended_version.Rmd
renamed paper versions

@@

  Here, we present a Bayesian hierarchical model that uses Poisson regression and state-space representation to produce absolute estimates of past population size and density. Using the  area North of the main ridge of the Swiss Alps in prehistoric times (7000–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone. The model's estimates of the credibility of its prediction, and the relative weight it affords to individual proxies through time, give further insights into the relative reliability of the evidence currently available for paleodemographic research. Our prediction of population development of the case study area accords well with the current understanding in the wider literature, but provides a more precise and higher-resolution estimate that is less sensitive to spurious fluctuations in the proxy data than existing approaches, especially the popular summed probability distribution of radiocarbon dates.

@@

  - We provide absolute estimates of population size and density on the area north of the Swiss Alpes in prehistoric times (7000–1000 BCE)

@@

Our case study targets the period between 7000–1000 BCE. The lower limit of this window was chosen to avoid the so-called 'Hallstatt plateau' on the Northern Hemisphere radiocarbon calibration curve, which would cause difficulties for the ^14^C proxy data. The upper limit was chosen to coincide with post-glacial changes in pollen spectra, before which the openness indicator is highly unlikely to reflect human influence before 7000 in the working area.

@@

### Colophon

>>>>

    keep_md: yes
    includes:
        in_header: preamble.tex

@@

  Here, we present a Bayesian hierarchical model that uses Poisson regression and state-space representation to produce absolute estimates of past population size and density. Using the  area North of the main ridge of the Swiss Alps in prehistoric times (–1000 BCE) as a case study, we show that combining multiple proxies (site counts, radiocarbon dates, dendrochronological dates, and landscape openness) produces a more robust reconstruction of population dynamics than any single proxy alone. The model's estimates of the credibility of its prediction, and the relative weight it affords to individual proxies through time, give further insights into the relative reliability of the evidence currently available for paleodemographic research. Our prediction of population development of the case study area accords well with the current understanding in the wider literature, but provides a more precise and higher-resolution estimate that is less sensitive to spurious fluctuations in the proxy data than existing approaches, especially the popular summed probability distribution of radiocarbon dates.

@@

  - We provide absolute estimates of population size and density on the area north of the Swiss Alpes in prehistoric times (–1000 BCE)

@@

Our case study targets the period between 6000–1000 BCE. The lower limit of this window was chosen to avoid the so-called 'Hallstatt plateau' on the Northern Hemisphere radiocarbon calibration curve, which would cause difficulties for the ^14^C proxy data. The upper limit was chosen to coincide with post-glacial changes in pollen spectra, before which the openness indicator is highly unlikely to reflect human influence before 7000 in the working area.

@@

# Author contributions
- *Martin Hinz*: Conceptualization, Methodology, Software, Validation, Formal analysis, Investigation, Data Curation, Writing - Original Draft, Writing - Review & Editing, Visualization
- *Joe Roe*: Software, Validation, Writing - Review & Editing
- *Julian Laabs*: Investigation, Data Curation, Writing - Review & Editing
- *Caroline Heitz*: Conceptualization, Investigation, Writing - Review & Editing
- *Jan Kolář*: Conceptualization, Writing - Review & Editing

# Colophon
